A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-06-15 tex/BiSig.tex
[ tex ] Comment out some unused packages
[ tex ] Change title
[ tex ] Add points to address; finalise the paper title
[ tex ] Using the same color as by Dunfield et al.
[ tex ] Minor
[ tex ] Draft the abstract

\documentclass[acmsmall,screen]{acmart}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[british]{babel}

\usepackage{bm,braket,mathtools,mathpartir,bussproofs}
\EnableBpAbbreviations
\usepackage[all,cmtip]{xy}

\usepackage{subcaption,xfrac,xspace,xcolor,xspace}
\usepackage{hyperref}
\usepackage[capitalise,noabbrev]{cleveref}
\usepackage[inline]{enumitem} % for environment enumerate*
\setlist[enumerate]{mode=unboxed}

\setlength {\marginparwidth }{2cm}
\usepackage[color=yellow,textsize=small]{todonotes}
\newcommand{\LT}[1]{\todo[author=LT,inline,color=green!40,caption={}]{{#1}}}
\newcommand{\Josh}[1]{\todo[author=Josh,inline,color=green!40,caption={}]{{#1}}}

@@

\citestyle{acmauthoryear}

\AtEndPreamble{%
\theoremstyle{acmdefinition}
\newtheorem{remark}[theorem]{Remark}}

@@

@@

\title{A Constructive Metatheory of Bidirectional Type Synthesis for Simple Types}

\begin{abstract}
  This paper proposes a formal theory of bidirectional typing.
\end{abstract}

>>>>

\documentclass[acmsmall,screen]{acmart}

\ifPDFTeX
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[british]{babel}
\else
\usepackage{polyglossia}
\setdefaultlanguage{british}
\fi

\usepackage{bm}
\usepackage{mathtools}
\usepackage{mathpartir}
\input{mathlig}
%\usepackage{xfrac}
\usepackage{xspace}
%\usepackage{hyperref}
\usepackage[capitalise,noabbrev]{cleveref}

%\usepackage[inline]{enumitem} % for environment enumerate*
%\setlist[enumerate]{mode=unboxed}

\setlength {\marginparwidth }{2cm}
\usepackage[color=yellow,textsize=small]{todonotes}

@@

\citestyle{acmauthoryear}
\AtEndPreamble{%
\theoremstyle{acmdefinition}
\newtheorem{remark}[theorem]{Remark}}

@@

\newcommand{\LT}[1]{\todo[author=LT,inline,color=green!40,caption={}]{{#1}}}
\newcommand{\Josh}[1]{\todo[author=Josh,inline,color=green!40,caption={}]{{#1}}}

@@

\title{A Theory of Bidirectional Type Synthesis for Simple Types}

\begin{abstract}
%  Type checking (and inference) can be seen as the transition from parsed abstract syntax trees to well-typed programs.
%  Many techniques and principles have been developed to type check advanced type theories and systems.
%  However, as theory and practice of parsing are able to address common grammars uniformly and parser generators have existed, type checkers by now are still designed and implemented on a case-by-case basis, not to mention a general theory of type checking and a robust `type checker generator' that can handle languages uniformly.
%
%  This paper is a step towards a \emph{general theory} of type checking and inference.
%  Inspired by recent language-formalisation frameworks, we theroise bidirectional typing---a design discipline of type synthesis mutually defining type checking and inference---for \emph{arbitrary} bidirectional type systems specified by a bidirectional binding signature and also formalise the theory in vanilla \Agda with Axiom K.
%  Prominently, our theorems including decidability of type synthesis are sound and complete type synthesisers which are correct by construction.
  Type checking and inference serve as the passage from parsed abstract syntax trees to well-typed programs.
  While significant progress has been made in developing techniques for type checking advanced type systems on a case-by-case basis, the lack of a general theory and a robust `type checker generator' poses a challenge to achieving uniformity across diverse programming languages.

  Motivated by this challenge, this paper contributes to the development of a general and practical theory of type checking and inference.
  We present a \emph{theory} of bidirectional typing, which had only been a \emph{design discipline}, for synthesising types to handle \emph{syntax-directed} bidirectional type systems of simple types that can be specified by a bidirectional binding signature.
  Here, \emph{type synthesis} refers to the mutual use of type checking and inference to synthesise a type from programs possibly with type annotations.

  To establish the correctness and effectiveness of our proposal, we formalize our theory in vanilla \Agda.
  We prove logical decidability of type synthesis, subsuming soundness and completeness of type synthesis algorithm.
  By Curry-Howard correspondence, our theorems are actually correct-by-construction synthesisers.
  We also discuss another essential characteristic of bidirectional typing called \emph{annotability} and propose a related but simpler and practical notion called \emph{bidirectionalisation}.
  Combining with decidability of type synthesis, we derive the trichotomy of raw terms by type synthesis.
  Put differently, the trichotomy is a type synthesiser which distinguishes programs with missing type annotations from programs that cannot be checked.

  In summary, this paper represents a step towards a general and practical theory of type checking and inference.
  By formalizing our theory of bidirectional type synthesise in \Agda, we provide a framework capable of handling bidirectional type systems of simple types.
  Our work addresses the problem with existing methodologies, setting the stage for further advances towards a general theory of type synthesis.
\end{abstract}

\keywords{\Agda, abstract syntax, bidirectional typing, type synthesis, type checking, type inference, language formalisation}

### 2023-06-15 tex/sec1-intro.tex
[ tex ] Added acmart files; separate compilation for TeXShop
[ tex ] Fix bib (Jones -> Peyton-Jones)
[ tex ] Draft the abstract

\begin{enumerate}
  \item Type checking as a transition from (parsed) abstract syntax trees to well-typed terms.
  \item Bidirectional typing is the only solution that works that scales well.
  \item Not yet a general theory but by now a design principle.
    No `type checker generator' as each type checker is still designed and implemented on a case-by-case basis, but can be derived from a syntax-directed type theory `easily'.
  \item The existing work of bidirectional typing does not take account of missing annotations carefully enough. 
  \item A meta-problem: existing work is based on (or at least left unspecified) classical logic.
    One has to state soundness of completeness of their type checking algorithm in a less straightforward way.
  \item A review of what has been done towards this line of research.
  \item Finally, our responses (contributions of this paper).
\end{enumerate}

@@

\cite{Pierce2000}
\cite{Dunfield2021}
\cite{Jones2007}

@@

Plan of this paper:
\begin{enumerate}
  \item Key ideas (\Cref{sec:key-ideas})
  \item Definitions for simple type theories and for bidirectional type systems (signature erasure and mode annotation) (\Cref{sec:defs})
  \item Soundness, completeness, bidirectionalisation, and annotatability (\Cref{sec:annotatability})
    \Josh{Erasure from bidirectional typing derivation to raw terms with a mode}
  \item Bidirectional type inference (\Cref{sec:type-synthesis})
  \item Formalisation with further examples (\Cref{sec:formalisation})
  \item Future work (\Cref{sec:future})
\end{enumerate}
  \LT{better title?}

>>>>

Points to elaborate:
\begin{enumerate}
  \item Type inference and checking as a transition from (parsed) abstract syntax trees to well-typed terms.
  \item Bidirectional typing is by now the only solution that scales well.
  \item Not yet a general theory but by now a design principle.
    No `type checker generator' as each type checker is implemented on a case-by-case basis, but an algorithm can usually be read off from syntax-directed typing rules.
    A hint to a general pattern in syntax-directedness.
  \item The existing work of bidirectional typing does not take account of practical aspects such as missing annotation.
    Lightly or rarely discussed.
  \item A meta-problem: existing work is based on classical logic (or left unspecified at least).
    Soundness of completeness of a type checking algorithm in a less straightforward way.
  \item A review of what has been done towards this line of research.
  \item Finally, our responses (contributions of this paper).
\end{enumerate}

@@

\cite{Pierce2000}
\cite{Peyton-Jones2007}
\cite{Dunfield2021}

@@

Plan of this paper:
\begin{enumerate}
  \item Key ideas (\Cref{sec:key-ideas})
  \item Definitions for bidirectional type systems (bidirectional binding signature, bidirectional type systems, signature erasure and mode annotation) (\Cref{sec:defs})
  \item Soundness, completeness, bidirectionalisation, and annotatability (\Cref{sec:annotatability})
    \Josh{Erasure from bidirectional typing derivation to raw terms with a mode}
  \item Bidirectional type inference (\Cref{sec:type-synthesis})
  \item Formalisation with further examples (\Cref{sec:formalisation})
  \item Future work (\Cref{sec:future})
\end{enumerate}
  \LT{better title?}
